---
title: "Gamete distance calculations"
author: "George Roff"
date: "2024-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Base parameterisation

see Gibbs (2024) Restoration Ecology for full methods

To calculate the encounter rate $E_{\text{turbulence}}$,use the formula:

$$
E_{\text{turbulence}} = 1.9 \times C \times \pi \times R^{\frac{7}{3}} \times \varepsilon^{\frac{1}{3}}
$$

where:

-    $E_{\text{turbulence}}$: Encounter rate (encounters per second)

-    $C$: Gamete density (eggs/m\^3)

-    $R$: Interactive distance (m)

-    $\varepsilon$: Turbulent dissipation rate (m²/s³)

-    1.9 = Lead constant

```{r, message=FALSE}

library(tidyverse)
library(ggplot2)

gamete_density <- 10000*100                # eggs/m^3
interactive_distance <- 1.5/1000       # mm to meters
epsilon <- 1e-4                        # m^2/s^3 (turbulent dissipation rate)
lead_constant <- 1.9                   # lead parameter constant


E_turbulence <- lead_constant * gamete_density * pi * interactive_distance^(7/3) * epsilon^(1/3)
paste0(round(E_turbulence*60, 4), " encounters per minute")






```

### Factorial parameterisation

For orthogonal parameterisation across different parameter levels (gamete density, interactive distance values, and dissipation rates):

```{r, width=12, height=12}

  # Define ranges for each parameter
gamete_density_values <- c(1e5, 1e6, 1e7)*100       # Gamete density (eggs/m^3)
interactive_distance_values <- seq(1, 5, 1)     # Interactive distance (mm)
epsilon_values <- c(1e-8, 1e-7, 1e-6, 1e-5, 1e-4)  # Turbulent dissipation rate (m^2/s^3)
lead_constant <- 1.9                            # Fixed lead parameter constant

# Create an orthogonal design (full factorial design) for the parameters
results <- expand.grid(
  gamete_density = gamete_density_values,
  interactive_distance_mm = interactive_distance_values,
  epsilon = epsilon_values
) |> mutate(lead_constant = lead_constant)


# Define the formula as a function
calculate_encounter_rate <- function(gamete_density, interactive_distance_mm, epsilon, lead_constant) {
  interactive_distance_m <- interactive_distance_mm / 1000   # Convert interactive distance from mm to meters
  encounter_rate <- lead_constant * gamete_density * pi * interactive_distance_m^(7/3) * epsilon^(1/3)   # Calculate the encounter rate based on the given formula
  return(encounter_rate)
}

# Apply the function across each row in the design data frame to calculate encounter rates
results$E_turbulence <- mapply(
  calculate_encounter_rate, 
  gamete_density = results$gamete_density,
  interactive_distance_mm = results$interactive_distance_mm,
  epsilon = results$epsilon,
  lead_constant = results$lead_constant
)


ggplot(results, aes(x = as.factor(gamete_density), y = as.factor(interactive_distance_mm), fill = (E_turbulence*60))) +
  geom_tile(color = "black") +
  facet_wrap(~ epsilon, scales = "free") +  # Facet by epsilon levels
  #scale_fill_viridis_c(option = "viridis") +  # Use a color scale for E_turbulence
  scale_fill_distiller(palette="RdBu", direction=-1) +  # Use a color scale for E_turbulence
  labs(
    x = "Gamete Density (eggs/m^3)",
    y = "Interactive Distance (mm)",
    fill = "Encounter Rate\n(E_turbulence)",
    title = "Encounter Rate (encounters per ",
    subtitle = "Faceted by Turbulent Dissipation Rate (epsilon)"
  ) +
  theme_minimal() +
  theme(
    legend.position="bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),   
    strip.text = element_text(size = 10, face = "bold") 
  )



```
